---
title: "SAS highlighting in Rmarkdown to PDF exporting"
author: "Tom Ratford"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Requirements

1. MikTex (or some equivalent)
2. A python environment (with Pygments installed via `pip install Pygments`)

Ensure both of these are setup and run as expected in RStudio.

## Pandoc

R convert `.Rmd` script to HTML/TeX via a program called `Pandoc`.
`Pandoc`'s syntax highlighter of choice does not support SAS.
There are a few articles on how to get SAS highlighting in R markdown files being knitted to HTML.

 * [R markdown syntax highlighting](https://www.calex.org/blog/r-markdown-syntax-highlighting/)
 * [Highlighting SAS chunks in rmarkdown](https://www.atorusresearch.com/highlighting-sas-chunks-in-rmarkdown/)

However these all require external javascript packages to run in HTML, and are not suitable with the TeX PDF output.

We could use the `highr` package with another external program called [highlight](http://www.andre-simon.de/doku/highlight/en/highlight.php).
This program can then be called to generate SAS code in LaTeX format.  
```{r, eval=FALSE}
highr::hi_andre(code = ..., 
                language = "sas",
                format = "latex")
```
We can then create a `knitr` engine (like we do in this document)

## Minted

`minted` is a LaTeX package which highlights code using a python lexer.
You can find a simple guide on [overleaf](https://www.overleaf.com/learn/latex/Code_Highlighting_with_minted#One-line_code), and the full documentation [here](https://texdoc.org/serve/minted.pdf/0)
We will use this to create a simple `knitr` engine to produce our highlighted SAS code.

## Project Setup

Optionally, ensure you are in a R project in RStudio, and create a `.Rprofile` file.
We have to change the arguments passed to the LaTeX engine to use shell escapes (as `minted` requires this).
We need the following R code to run:
```{r, eval=FALSE}
options(tinytex.engine_args="-shell-escape")
```
In every session. You can achieve this by either:
 * Putting this into your `.Rprofile` file
 * Running this every session

In your R markdown header put the following
```{r, eval=FALSE}
---
title: 'My Doc'
author: "Tom Ratford"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - \usepackage{minted}
  - \usemintedstyle{tango}
---
```

## Knitr engine

In your R setup put the following
```{r, eval=FALSE}
knitr::knit_engines$set(sas = function(options) {
  code <- paste("\\begin{minted}[bgcolor=shadecolor]{sas}",
                paste(options$code, collapse = "\n"),
                "\\end{minted}",
                sep="\n")
})
```

you will now be able to write the following to get highlighted SAS code in your pdf output.
```{}
\```{sas}
 put sas code here
\```
```

## Alterative header options

If you have many options and packages you wish to run, I recommend using this format.
```{r, eval=FALSE}
---
title: 'mydoc'
author: "Tom Ratford"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    extra_dependencies: ["minted", ...]
    includes:
      in_header : preamble.tex
      before_body: doc-prefix.tex
      after_body: doc-suffix.tex
---
```
If you have other LaTeX packages you wish to use, put them in the `...` box, else delete this section.
The `.tex` files must be in the the project root directory. 

## Themes
There are many [themes supported in .Rmd blocks](https://bookdown.org/yihui/rmarkdown/html-document.html#appearance-and-style) as well as in [minted blocks](https://www.overleaf.com/learn/latex/Code_Highlighting_with_minted#Reference_guide). 
However there is little overlap. 
The overlap (I have found) is:

 * BOTH = tango
 * Minted = default, .Rmd = pygments
 * Minted = Xcode, .Rmd = textmate
 
`minted` also supports a sas theme to make the code look like it does in the SAS program editor.